---
title: "Spillover estimation IMC"
author: 'Vito Zanotelli et al.'
output:
  html_document:
    df_print: paged
    keep_md: true
---

# Aim
This script shows how to estimate spillover from single metal spots on an agarose coated slide.
This short script illustrates how the helper function 'estimate_sm_from_imc_txtfol' can be used to estimate a spillover matrix from a folder containing all acquisitions 
in .txt format in one single function call.

```{r Libraries, message=FALSE}
library(CATALYST)
library(data.table)
library(ggplot2)
library(flowCore)
library(dplyr)
library(dtplyr)
library(stringr)
library(ggpmisc)
source('spillover_imc_helpers.R')
```


## setup the configuration variables
```{r Setup}
# a folder containing a complete single stain acquisition
fol_ss = '../data/Figure_S5/Spillover_Matrix_2'
```
Example for estimation of the spillover matrix from a folder of single stains
with single stains parsed from naming: xxx_x_metal_x.txt? (E.g. Dy161 1-1000_8_Dy161_8.txt)
```{r}
#' estimate_sm_from_imc_txtfol
#' Estimates spillover directly from a folder containing IMC .txt files
#'
#' @param fol_ss folder containing .txt acquisitions of IMC single stains
#' @param ssmetals_from_fn logical, Are the single stains correctly named xxx_x_metal_x.txt? (E.g. Dy161 1-1000_8_Dy161_8.txt
#' @param ssmetals Vector of masses of the single stains used. Required if ssmetals_from_file_fn is False
#' @param fn2ssmetal Optional: a named vector mapping the filenames to the single stain metal used (e.g. if it cannot be parsed from the filename)
#' @param remove_incorrect_bc Remove barcodes not matching the filename single stain annotation (requires either ssmetals_from_fn=T or fn2ssmetal )
#' @param minevents Minimal number of events (after debarcoding) that need to be present in a single stain in order that a spillover estimation is performed
#' @param bin_n_pixels Optional: integer, bin n consecutive pixels. Can be used if the intensities per pixel are to low (e.g. <200 counts)
#' @param ... Optional parameters will be passed to CATALYST::computeSpillmat 
res = estimate_sm_from_imc_txtfol(fol_ss, ssmetals_from_fn=T)
```
Plot of the spillover matrix estimated
(1:10000 is just to circumvent the requirement of the plotSpillmat function to have single stain masses provided)
```{r}
CATALYST::plotSpillmat(1:10000,res[['sm']])
```

Quality control: Distribution of medians per file

Note the fig.height and fig.width parameters, that can be adjusted if your own data has more facets.
```{r fig.height=20, fig.width=20}
plot_file_medians(res[['data']])
```

Additionally transforming the y axis with a squareroot transform can make the plots more readable, especially if outlier channels are acquired.
```{r, fig.height=20, fig.width=20}
p = plot_file_medians(res[['data']])
p + scale_y_sqrt()
```

In some cases, e.g. when Argon was acquired, the plot above might not be informative, thus one can also calculate the medians directly and either write them to csv or use the Rstudio variable browser to inspect the results, e.g.
```{r}
dat_medians = calc_file_medians(res[['data']]) %>%
  dcast('metal+mass+file ~ variable', value.var='med')
dat_medians
```


In case the spillover matrix should be written out
```{r}
write.csv(res[['sm']],file = '../data/sm.csv')
```


Alternatively the function can also be called with a list of single stain masses, if the metals are not contained in the filename
```{r}
ssmass = c(161, 162, 163, 164, 166, 167, 168, 170, 151, 153, 155, 156, 158, 160, 165, 113, 115, 175, 142, 143, 144, 145, 146, 148, 150, 141, 147, 149, 152, 154, 159, 169, 171, 172, 173, 174, 176)
res = estimate_sm_from_imc_txtfol(fol_ss, ssmetals_from_fn=F,ssmass = ssmass, remove_incorrect_bc = F )
```
```{r}
CATALYST::plotSpillmat(1:10000,res[['sm']])
```


```{r}
sessionInfo()
```

