---
title: "Comparison of spill estimation methods"
author: "Helena L Crowell"
output: html_document
---

# Aim

This script investigates and compares different methods for spillover estimation and compensation:

* single-cell vs. population based estimation
* consideration of expected vs. all interactions
* classic vs. NNLS compensation

```{r warning=FALSE}
# load required packages
library(CATALYST)
library(flowCore)
library(ggplot2)
library(gridExtra)
library(matrixStats)
library(reshape2)

# source helpers
source('plotting-wrappers.R')
```

### Population vs. single-cell based estimation

```{r message=FALSE, warning=FALSE, fig.width=4}
# read in single-stained breads 
fcs_path <- '../data/Figure_S3/bead replicates/160805_beads.fcs'
ff <- read.FCS(fcs_path)

# assign preliminary IDs
ss_ms <- c(139, 141:156, 158:176)
res <- assignPrelim(ff, ss_ms, verbose=FALSE)

# apply deconvolution parameters
sep_cutoffs <- read.table('../data/Figure_S3/sep_cutoffs.txt', check.names=FALSE)
cutoffs <- c(.4, sep_cutoffs$`160805`)
res <- applyCutoffs(res, mhl_cutoff=20, cutoffs)

# estimate spillover matrices & compensate
sm_classic <- computeSpillmat(res, method="classic", trim=.01)
sm_default <- computeSpillmat(res, method="default")
sms <- list(sm_classic, sm_default)

comped_classic <- compCytof(ff, sm_classic)
comped_default <- compCytof(ff, sm_default)

# get indices of channels expected to receive spillover
chs <- colnames(ff)
ms <- get_ms(chs)
mets <- get_mets(chs)
spill_cols <- CATALYST:::get_spill_cols(ms, mets)

# helper to get medians
get_meds <- function(ff) {
    unlist(sapply(ss_ms, function(i) {
        inds <- bc_ids(res) == i
        cols <- spill_cols[[which(ms == i)]]
        cols <- cols[ms[cols] %in% 138:176]
        colMedians(as.matrix(exprs(ff[inds, cols])))
    }))
}
meds_uncomped <- get_meds(ff)
meds_comped_classic <- get_meds(comped_classic)
meds_comped_default <- get_meds(comped_default)
df <- data.frame(meds_uncomped, meds_comped_classic, meds_comped_default)
colnames(df) <- c("Uncompensated", 
    "Population based\nestimation", 
    "Single-cell based\nestimation")
df <- melt(df)
plot_method_comparison(df)
```

```{r warning=FALSE, fig.width=8, fig.height=6}
# (for convenience, we will use the same set of beads as above)
# get spill values for interactions in range of interest
idx <- which(ms %in% ss_ms)
from <- rep(idx, sapply(spill_cols[idx], length))
to <- unlist(spill_cols[idx])

spill_vals <- 100 * matrix(do.call(cbind, lapply(sms, function(i) 
    CATALYST:::make_symetric(i)[cbind(from, to)])), ncol=length(sms))

# plot only interactions above threshold
fil <- rowMeans(spill_vals) > .5
spill_vals <- spill_vals[fil, ]
# order increasingly
o <- order(rowMeans(spill_vals))
spill_vals <- spill_vals[o, ]
deltas <- apply(spill_vals, 1, diff) 

# get labels for interactions
from <- chs[from[fil][o]]
to <- chs[to[fil][o]]
rownames(spill_vals) <- factor(paste0(from, "->", to))
colnames(spill_vals) <- c("Population based estimation", "Single-cell based estimation")

df1 <- reshape2::melt(spill_vals) 
df2 <- data.frame(x=seq_len(sum(fil)), y=deltas)
df3 <- data.frame(x=seq_len(sum(fil)), y=100*deltas/rowMeans(spill_vals))
plot_classic_vs_default(df1, df2, df3)
```

### Expected crosstalk interactions

The code snippet below highlights the interactions taken into consideration by the default method for spillover estimation with `computeSpillmat` in `CATALYST`.

```{r message=FALSE, fig.width=4}
# read in examplary single-stains
data(ss_exp)

# for each channel, get indices of channels 
# that are expected to receive spillover
chs <- colnames(ss_exp)
ms <- get_ms(chs)
mets <- get_mets(chs)
spill_cols <- CATALYST:::get_spill_cols(ms, mets)

# initialize empty spillover matrix
sm <- diag(length(ms))
rownames(sm) <- colnames(sm) <- chs

# fill in some value where spillover is expected
ss_ms <- c(139, 141:156, 158:176)
for (m in ss_ms) {
    idx <- which(ms == m)
    sm[idx, spill_cols[[idx]]] <- .5
}

# plot spillover matrix
plotSpillmat(ss_ms, sm, annotate=FALSE)
```

### Comparison of 4 estimation modes

```{r fig.width=4.5}
# deconvolution of single-stained beads
fcs_path <- '../data/Figure_S3/bead replicates/160616_beads.fcs'
ff <- read.FCS(fcs_path)
ss_ms <- c(139, 141:156, 158:176)
res <- assignPrelim(ff, ss_ms, verbose=FALSE)
sep_cutoffs <- read.table('../data/Figure_S3/sep_cutoffs.txt', check.names=FALSE)
cutoffs <- c(.4, sep_cutoffs$`160616`)
res <- applyCutoffs(res, mhl_cutoff=20, cutoffs)

# estimate SM with all modes (note: for method="classic", 
# the trim-value was selected manually via CATALYST::estTrim)
sm_default     <- computeSpillmat(res)
sm_default_all <- computeSpillmat(res, interactions="all")
sm_classic     <- computeSpillmat(res, method="classic", trim=.01)
sm_classic_all <- computeSpillmat(res, method="classic", interactions="all", trim=.01)
sms <- list(sm_default, sm_default_all, sm_classic, sm_classic_all)

# keep only range of interest
chs <- colnames(sms[[1]])
ms <- get_ms(chs)
keep <- ms %in% 139:176
sms <- lapply(sms, function(sm) sm[, keep])
# get highest off-diagonal spill
norm_fact <- max(sapply(sms, function(sm) {
    sm <- CATALYST:::make_symetric(sm)
    max(sm[row(sm) != col(sm)])
}))
labs1 <- paste("method =", rep(c("\"default\"", "\"classic\""), each=2))
labs2 <- paste("interactions =", rep(c("\"default\"", "\"all\""), 2))
mains <- paste0(paste(labs1, "\n"), labs2)
# wrapper plotSpillmat2 shows spill < 0.001% in black,
# and scales spill values from 0-1 for visualization
for (i in seq_along(sms))
    plotSpillmat2(ss_ms, sms[[i]], norm_fact, mains[i])
```